---
title: "Air Quality Data Exploration"
author: "Brian High"
date: "Thursday, May 07, 2015"
output: 
    html_document:
        keep_md: yes
---

Setup
-----

Load the required packages.

```{r, echo=TRUE, message=FALSE}
for (pkg in c("dplyr", "reshape2", "ggplot2")) {
    if (! require(pkg, character.only=TRUE)) { 
        install.packages(pkg, repos="http://cran.fhcrc.org", dependencies=TRUE)
        suppressPackageStartupMessages(library(pkg, character.only=TRUE))
    }
}
```

Load the dataset
----------------

Let's explore the [New York Air Quality Measurements](http://stat.ethz.ch/R-manual/R-patched/library/datasets/html/airquality.html) from R's `dataset` package. We'll start
by loading the dataset and examining the structure.

```{r}
library(datasets)
data(airquality)

str(airquality)
```

View some rows
--------------

It is always a good idea to take a look at a few rows of data to get oriented.

```{r}
head(airquality)
tail(airquality)
```

Basic statistical summary
-------------------------

With `summary`, we get a quick view of `NA`s, max, min, median, mean and first 
and third quartiles.

```{r}
summary(airquality)
```

Pairwise matrix of scatterplots
-------------------------------

Using `pairs`, we can examine relationships between pairs of variables.

```{r}
pairs(airquality, panel = panel.smooth)
```

Linear model fit
----------------

We can produce a summary of linear model fit statistics with `lm` and `summary`.

```{r}
summary(lm(Ozone ~ Wind + Solar.R + Temp, data = airquality))
```

Remove the NA values
--------------------

To avoid warnings with `qplot` and `ggplot`, we first need to remove the `NA` 
values with `na.omit`. We will create a new `data.frame` in the process, so that
the original dataset will remain unchanged.

```{r}
df <- na.omit(as.data.frame(airquality))
summary(df)
```

Plot Ozone versus Temp
----------------------

Plot `Ozone` versus `Temp` with a linear regression fit using `qplot` and `method="lm"`.

```{r}
qplot(Temp, Ozone, data=df, geom=c("point", "smooth"), method="lm")
```

Add a Date column
-----------------

Knowing the data was collected in 1973, we can combine this year with the `Month` 
and `Day` to get a `Date` column. That will allow us to produce a time series plot.

```{r}
df$Date <- as.Date(paste("1973", df$Month, df$Day, sep="-"), "%Y-%m-%d")
str(df)
```

Reshape Temp and Ozone
----------------------

We will `select` only the `Temp`, `Ozone`, and `Date` variables, then "reshape"
the dataset into a "long" format with `melt`.

```{r}
df <- select(df, Temp, Ozone, Date) 
df <- melt(df, id="Date")

str(df)
```

Time Series Plot
----------------

Now we can plot `Temp` and `Ozone` in a time series, with "loess" curves. 

```{r}
ggplot(df, aes(x=Date, y=value, colour=variable, group=variable)) +
   geom_point(aes(y=value, colour=variable)) + geom_smooth(method="loess")
```